---
title: "SI: Settlements"
author:
  - Petr Pajdla:
      email: petr.pajdla@protonmail.com
      institute: [UofO]
      correspondence: false
  - František Trampota:
      email: trampota@phil.muni.cz
      institute: [UofO]
      correspondence: true
institute:
  - UofO: Department of Archaeology and Museology, Masaryk University, Brno
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl"
# abstract: |
keywords: |
  keyword 1; keyword 2; keyword 3
editor_options: 
  markdown: 
    wrap: 80
---

Keywords: `r rmarkdown::metadata$keywords`

```{r setup, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  cache = TRUE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

```{r packages}
devtools::load_all(".")
library(sf)
library(tidyverse)
```

# Introduction

The analysis is divided into two 'time' scales:

-   Pottery traditions
-   Pottery groups

The study area is broadly defined as *Easter part of Bohemia* (abbreviated *B*)
and *Morava river catchment* (*M*). It spans across the area of the Czech
Republic and bordering regions of Austria and Slovakia.

<!-- ======================================================================= -->

# Background

```{r get-data, cache = FALSE}
derived_data <- "analysis/data/derived_data"
path_maps <- here::here(derived_data, "maps/")
path_data <- here::here(derived_data)
path_temp <- here::here(derived_data, "temp/")

# mask
mask <- st_read(paste0(path_maps, "mask.geojson"), quiet = TRUE)
bbox <- st_read(paste0(path_maps, "bbox.geojson"), quiet = TRUE)

# rivers + admin for plots
admin0 <- st_read(paste0(path_maps, "admin0.geojson"), quiet = TRUE)
admin1 <- st_read(paste0(path_maps, "admin1.geojson"), quiet = TRUE)
rivers0 <- st_read(paste0(path_maps, "rivers0.geojson"), quiet = TRUE)
rivers1 <- st_read(paste0(path_maps, "rivers1.geojson"), quiet = TRUE)

# labels
labs_chrono <- read_rds(here::here(derived_data, "chrono_labels.RDS"))

# database
set_base <- read_rds(here::here(derived_data, "settlements.RDS"))

set_spat <- st_read(here::here(derived_data, "settlements_sf.geojson"), 
                    quiet = TRUE)

settlements1 <- set_spat %>%
  right_join(set_base$period1, by = "id")
settlements2 <- set_spat %>%
  right_join(set_base$period2, by = "id")
```

## Pottery traditions

```{r tab-set1-count}
set_base$period1 %>%
  group_by(period_label) %>%
  summarize(n = n()) %>% 
  rename('Period' = period_label) %>% 
  knitr::kable(caption = "Number of settlements across time periods.") %>% 
  kableExtra::kable_styling(full_width = FALSE)
```

```{r plot-set1-count, fig.cap = 'Settlements classified by general pottery traditions across time periods.'}
settlements1 %>%
  ggplot() +
  geom_sf(data = mask, fill = "gray80", color = NA, alpha = 0.4) +
  # geom_sf(data = admin1, linetype = 3, fill = NA, color = "gray60", size = 0.2, alpha = 0.4) +
  geom_sf(data = admin0, linetype = 1, fill = NA, color = "gray60", alpha = 0.4) +
  geom_sf(data = rivers1, color = "mediumblue", size = 0.4, alpha = 0.4) +
  geom_sf(data = rivers0, color = "mediumblue", size = 0.6, alpha = 0.4) +
  geom_sf(alpha = 0.2, aes(color = label)) +
  scale_color_brewer(palette = "Set1", name = "Tradition") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  facet_wrap(vars(period_label), nrow = 3) +
  theme_void() +
  theme(panel.border = element_rect(color = "black", fill = NA))
```

<!-- ----------------------------------------------------------------------- -->

## Pottery groups

```{r tab-set2-count}
set_base$period2 %>%
  group_by(period_label) %>%
  summarize(n = n()) %>% 
  rename('Period' = period_label) %>% 
  knitr::kable(caption = "Number of settlements across time periods.") %>% 
  kableExtra::kable_styling(full_width = FALSE)
```

```{r plot-set2-count, fig.cap = 'Settlements classified by pottery groups across time periods.'}
settlements2 %>%
  ggplot() +
  geom_sf(data = mask, fill = "gray80", color = NA, alpha = 0.4) +
  # geom_sf(data = admin1, linetype = 3, fill = NA, color = "gray60", size = 0.2, alpha = 0.4) +
  geom_sf(data = admin0, linetype = 1, fill = NA, color = "gray60", alpha = 0.4) +
  geom_sf(data = rivers1, color = "mediumblue", size = 0.4, alpha = 0.4) +
  geom_sf(data = rivers0, color = "mediumblue", size = 0.6, alpha = 0.4) +
  geom_sf(alpha = 0.2, aes(color = label)) +
  scale_color_brewer(palette = "Set1", name = "Tradition") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  facet_wrap(vars(period_label), nrow = 3) +
  theme_void() +
  theme(panel.border = element_rect(color = "black", fill = NA))
```

<!-- ======================================================================= -->

# Settlement density

Settlement density is estimated using KDE. Kernel size of 4 km is used. This
value is completely arbitrary but is selected because settlement clusters are
nicely highlighted at this scale.

## Pottery traditions

```{r set1-kde}
kde1 <- kdestimate(settlements1,
                   mask,
                   markscol = "period",
                   lvls = names(labs_chrono$period)) %>%
  rename(period = name) %>%
  mutate(period_label = labs_chrono$periods[period],
         period_label = factor(period_label, levels = labs_chrono$periods)) %>%
  select(id, starts_with("period"), chrono, kde)

# tibble with results
res1_long <- kde1 %>% 
  rename(settlements_kde = kde) %>% 
  select(-period_label) %>% 
  pivot_longer(col = settlements_kde, names_to = "variable")
```

```{r plot-set1-kde, fig.cap = 'Kernel density estimation for different periods.'}
kdestimate(settlements1,
           mask,
           markscol = "period",
           lvls = names(labs_chrono$period),
           points = FALSE) %>%
  mutate(period_label = labs_chrono$periods[period],
         period_label = factor(period_label, levels = labs_chrono$periods)) %>%
  ggplot(aes(x, y, fill = kde)) +
  geom_raster() +
  scale_fill_viridis_c(name = "KDE") +
  facet_wrap(vars(period_label)) +
  coord_fixed() +
  theme_void()
```

<!-- ----------------------------------------------------------------------- -->

## Pottery groups

```{r set2-kde}
kde2 <- kdestimate(settlements2,
                   mask,
                   markscol = "period",
                   lvls = names(labs_chrono$period)) %>%
  rename(period = name) %>%
  mutate(period_label = labs_chrono$periods[period],
         period_label = factor(period_label, levels = labs_chrono$periods)) %>%
  select(id, starts_with("period"), chrono, kde)

# tibble with results
res2_long <- kde2 %>% 
  rename(settlements_kde = kde) %>% 
  select(-period_label) %>% 
  pivot_longer(col = settlements_kde, names_to = "variable")
```

```{r plot-set2-kde, fig.cap = 'Kernel density estimation for different periods.'}
kdestimate(settlements2,
           mask,
           markscol = "period",
           lvls = names(labs_chrono$period),
           points = FALSE) %>%
  mutate(period_label = labs_chrono$periods[period],
         period_label = factor(period_label, levels = labs_chrono$periods)) %>%
  ggplot(aes(x, y, fill = kde)) +
  geom_raster() +
  scale_fill_viridis_c(name = "KDE") +
  facet_wrap(vars(period_label)) +
  coord_fixed() +
  theme_void()
```

<!-- ======================================================================= -->

# Distance to raw material sources

## Raw materials distribution

Raw material sources are defined as either points or lines. Each raw material
can be identified by single or multiple simple features. If the location of
source used in the Neolithic period is unknown or several sources or even modes
of procurement (e. g. from river bed etc.) are possible, the most probable are
included.

Minimum, mean, median and maximum distances are calculated, median is used 
as a result.

```{r get_data-rm}
rm_pts <- st_read(here::here(derived_data, "rm_points.geojson"), quiet = TRUE) %>% 
  mutate(label = if_else(orig == "l", "Chipped", "Polished"))
rm_lns <- st_read(here::here(derived_data, "rm_lines.geojson"), quiet = TRUE) %>% 
  mutate(label = if_else(orig == "l", "Chipped", "Polished"))
```

```{r plot-rm, fig.cap = "Distribution of raw material sources."}
ggplot() + 
  geom_sf(data = mask, fill = "gray80", color = NA, alpha = 0.4) +
  geom_sf(data = rivers1, color = "gray60", size = 0.4, alpha = 0.4) +
  geom_sf(data = rivers0, color = "gray60", size = 0.6, alpha = 0.4) +
  geom_sf(data = set_spat, alpha = 0.1, color = "gray80") +
  geom_sf(data = rm_pts, aes(color = rm), shape = 4) +
  geom_sf(data = rm_lns, aes(color = rm)) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  facet_wrap(vars(label)) +
  theme_void() +
  theme(panel.border = element_rect(color = "black", fill = NA)) +
  labs(color = "Raw material")
```

```{r rm-dist, cache = FALSE}
# calculated separately in a script `rm_distance.R`
# dist_pts <- dist_to_rm(set_spat, rm_pts, "rm")
# dist_lns <- dist_to_rm(set_spat, rm_lns, "rm")

rm_dist <- read_csv(paste0(path_temp, "rm_dist.csv"))

rm_dist_median <- rm_dist %>% 
  select(-region, -type, -shp, -min, -mean, -max) %>% 
  group_by(id, rm) %>% 
  summarise(dist_rm = mean(median)) %>% 
  pivot_longer(cols = dist_rm, names_to = "variable") %>% 
  mutate(variable = str_c(variable, rm, sep = "_")) %>% 
  select(-rm)

# update result tibbles
res1_long <- set_base$period1 %>% 
  select(id, period, chrono) %>% 
  inner_join(rm_dist_median) %>%
  bind_rows(res1_long)

res2_long <- set_base$period2 %>% 
  select(id, period, chrono) %>% 
  inner_join(rm_dist_median) %>%
  bind_rows(res2_long)
```

```{r plot-rm-dist-min, fig.cap="Distance to raw material sources."}
rm_dist %>%
  ggplot(aes(x = min, y = rm, fill = region)) +
  geom_violin() +
  scale_fill_brewer(palette = "Accent") +
  facet_wrap(~type, scales = "free_y") +
  labs(x = "min. dist. (km)", y = "raw material", fill = "Region") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r plot-rm-dist-kde-chrono1, fig.cap="KDE against distance to raw material source, Morava river catchment, pottery traditions."}
kde1 %>% select(id, period_label, kde) %>% 
  left_join(rm_dist) %>% 
  mutate(region = str_extract(id, "^.")) %>% 
  filter(rm %in% c("AD", "ZELESICE", "PMD", "KL"),
         region == "M") %>% 
  ggplot(aes(median, kde)) +
  geom_point(alpha = 0.1) + 
  stat_smooth(method = "lm", se = FALSE) +
  facet_grid(vars(period_label), vars(rm)) +
  labs(x = "Median distance to raw material source (km)",
       y = "KDE") +
  theme_minimal() +
  theme(strip.text.y = element_text(angle = 0))
```

```{r plot-rm-dist-kde-chrono2, fig.cap="KDE against distance to raw material source, Morava river catchment, pottery groups."}
kde2 %>% select(id, period_label, kde) %>% 
  left_join(rm_dist) %>% 
  mutate(region = str_extract(id, "^.")) %>% 
  filter(rm %in% c("AD", "ZELESICE", "PMD", "KL"),
         region == "M") %>% 
  ggplot(aes(median, kde)) +
  geom_point(alpha = 0.1) + 
  stat_smooth(method = "lm", se = FALSE) +
  facet_grid(vars(period_label), vars(rm)) +
  labs(x = "Median distance to raw material source (km)",
       y = "KDE") +
  theme_minimal() +
  theme(strip.text.y = element_text(angle = 0))
```

<!-- ======================================================================= -->

# Density of watercourses

In settlement structure studies it is often assumed that past human settlements
have some kind of a relation to the network of watercourses. Numerous problems
are faced while assessing the relationship between settlements and close
watercourse. Most importantly the man-made recent changes in the structure and
intensity of watercourses (and landscape in general) are a major factor
imapiring such analysis.

Usually, the settlement - watercourse relationship is expressed in some form of
a metric giving shortest distance or so. We choose an approach used to study
road networks in modern urban areas [e. g. @lin2020].

Most of the input data is derived from *DIBAVOD* data set A01 layer
(<https://www.dibavod.cz/>). 
Data for Slovakia are derived from OpenStreet map ? , data for Austria ???

Density (kernel smoothed intensity) is estimated using `spatstat::density.psp`
function with sigma 2000 m.

```{r get-raster}
hydro_dens <- raster::raster(
  here::here(derived_data, "hydro", "density_raster.tif")) %>% 
  raster::rasterToPoints() %>%
  as_tibble() %>%
  rename(KDE = density_raster)
```

```{r plot-hydro-density, fig.cap="Kernel smoothed intensity of watercourses overlaid with sites."}
ggplot() +
  geom_raster(data = hydro_dens, aes(x, y, fill = KDE)) +
  geom_sf(data = set_spat, alpha = 0.1, shape = 1, color = "white") +
  theme_void()
```

# Results

```{r results-export}
res1_long %>% 
  pivot_wider(names_from = variable, values_from = value) %>% 
  arrange(id, period, chrono) %>% 
  write_csv(here::here(derived_data, "results", "results1.csv"))

res2_long %>% 
  pivot_wider(names_from = variable, values_from = value) %>% 
  arrange(id, period, chrono) %>% 
  write_csv(here::here(derived_data, "results", "results2.csv"))
```


<!-- The following line inserts a page break -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational
environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```
